---
title: 'Preparing R: updates and installations'
author: "UC Davis Bioinformatics Core"
date: "`r Sys.Date()`"
always_allow_html: true
output:
  md_document:
    variant: gfm
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

For this course, you will need an up-to-date version of R (4.4.0 or higher) and RStudio. In the course of the workshop, we will be using a number of packages. Let's get them all installed before we start.

# Set up and installations

## BiocManager

```{r BiocManager}
if (!any(rownames(installed.packages()) == "BiocManager")){
  install.packages("BiocManager")
}
```

## Rmarkdown

```{r Rmarkdown}
if (!any(rownames(installed.packages()) == "rmarkdown")){
  BiocManager::install("rmarkdown")
}
```

# Cleaning and managing data

## lattice

```{r lattice}
if (!any(rownames(installed.packages()) == "dplyr")){
  install.packages("lattice")
}
```

## reshape2

```{r reshape2}
if (!any(rownames(installed.packages()) == "dplyr")){
  install.packages("reshape2")
}
```

## dplyr

```{r dplyr}
if (!any(rownames(installed.packages()) == "dplyr")){
  BiocManager::install("dplyr")
}
```

## tidyr

```{r tidyr}
if (!any(rownames(installed.packages()) == "tidyr")){
  BiocManager::install("tidyr")
}
```

## magrittr

```{r magrittr}
if (!any(rownames(installed.packages()) == "magrittr")){
  BiocManager::install("magrittr")
}
```

## kableExtra

```{r kableExtra}
if (!any(rownames(installed.packages()) == "kableExtra")){
  BiocManager::install("kableExtra")
}
```

# Basic visualizations

## ggplot2

```{r ggplot2}
if (!any(rownames(installed.packages()) == "ggplot2")){
  BiocManager::install("ggplot2")
}
```

## viridis

```{r viridis}
if (!any(rownames(installed.packages()) == "viridis")){
  BiocManager::install("viridis")
}
```

## ggExtra

```{r ggExtra}
if (!any(rownames(installed.packages()) == "ggExtra")){
  BiocManager::install("ggExtra")
}
```

## ggsignif

```{r ggsignif}
if (!any(rownames(installed.packages()) == "ggsignif")){
  BiocManager::install("ggsignif")
}
```

## ggrepel

```{r ggrepel}
if (!any(rownames(installed.packages()) == "ggrepel")){
  BiocManager::install("ggrepel")
}
```

## ggalluvial

```{r ggalluvial}
if (!any(rownames(installed.packages()) == "ggalluvial")){
  BiocManager::install("ggalluvial")
}
```

# Advanced visualizations

## circlize

```{r circlize}
if (!any(rownames(installed.packages()) == "circlize")){
  BiocManager::install("circlize")
}
```

## Signac

```{r Signac}
if (!any(rownames(installed.packages()) == "Signac")){
  BiocManager::install("Signac")
}
```


## Seurat

```{r Seurat}
if (!any(rownames(installed.packages()) == "Seurat")){
  BiocManager::install("Seurat")
}
```

## ggtree

```{r ggtree}
if (!any(rownames(installed.packages()) == "ggtree")){
  BiocManager::install("ggtree")
}
```

## ggcoverage

```{r ggcoverage}
if (!any(rownames(installed.packages()) == "ggcoverage")){
  install.package("remotes")
  remotes::install_github("showteeth/ggcoverage")
}
```

## Complex Heatmap

```{r ComplexHeatmap}
if (!any(rownames(installed.packages()) == "ComplexHeatmap")){
  BiocManager::install("ComplexHeatmap")
}
```

# Download markdown documentation

```{r download.file}
download.file("https://raw.githubusercontent.com/ucdavis-bioinformatics-training/2025-August-Intermediate-Visualization-for-Bioinformatics/R/01-boxplot.Rmd")
```

